Verification of Stencil Computations through Spatial Specifications

### 2016-07-04 paper.tex
add further equations and expanded the description; start on the model section

\newcommand{\sten}[1]{\textcolor{darkdarkgreen}{\!#1}}

@@

\newcommand{\stenRefl}[1]{\term{reflexive} \, (\term{dim=}#1)}
\newcommand{\stenFwd}[2]{\term{forward} \, (\term{depth=}#1,
  \term{dim=}#2)}
\newcommand{\stenBwd}[2]{\term{backward} \, (\term{depth=}#1,
  \term{dim=}#2)}
\newcommand{\stenCen}[2]{\term{centered} \, (\term{depth=}#1,
  \term{dim=}#2)}
\newcommand{\stenReflS}[1]{\term{refl} \, (\term{dim=}#1)}
\newcommand{\stenFwdS}[2]{\term{fwd} \, (\term{depth=}#1,
  \term{dim=}#2)}
\newcommand{\stenBwdS}[2]{\term{bwd} \, (\term{depth=}#1,
  \term{dim=}#2)}
\newcommand{\stenCenS}[2]{\term{cen} \, (\term{depth=}#1,
  \term{dim=}#2)}

@@

\begin{figure}[t]
\begin{align*}
\def\arraystretch{1.2}
\setlength{\arraycolsep}{0.2em}
\newcommand{\dimTy}{\mathbb{D}}
\begin{array}{rl}
\nonterm{specification} ::= & \nonterm{regionDec} \mid \nonterm{specDec} \\
\nonterm{specDec} ::= & \term{stencil} \; \nonterm{specTop} \;
                        \texttt{::} \; v
  \\
\nonterm{regionDec} ::= &  \texttt{region} \; \nonterm{rvar} \; \texttt{=} \;
                         \nonterm{region}\\[1em]
%\nonterm{spec} ::= & \nonterm{spatial} \mid \nonterm{temporal}
%\\[1em]
\nonterm{specTop} ::= & [\nonterm{approxMod},] \; \nonterm{spec} \\
\nonterm{spec} ::= & [\nonterm{modMult},] \; \nonterm{refRegion} \\
\nonterm{refRegion} ::= & [\nonterm{modRefl},] \; \nonterm{region}\\
\nonterm{modMult} ::= & \term{readOnce} \\
\nonterm{modRefl} ::= & \term{irreflexive} \, (\term{dims=}\dimTy \{\term{,}
          \dimTy\}) \\
\nonterm{approxMod} ::= & \term{atMost} \; \mid \; \term{atLeast} \\[0.1em]
\nonterm{region} ::=
        & \stenFwd{\mathbb{N}_{>0}}{\dimTy}  \\
\mid \; & \stenBwd{\mathbb{N}_{>0}}{\dimTy} \\
\mid \; & \stenCen{\mathbb{N}_{>0}}{\dimTy} \\
\mid \; & \stenRefl{\dimTy}  \\
\mid \; & \nonterm{region} \; \term{+} \; \nonterm{region} \\
\mid \; & \nonterm{region} \; \term{*} \; \nonterm{region} \\
\mid \; & \nonterm{rvar}  \\[0.5em]
%\nonterm{temporal} ::= \; & \term{dependency} \; (v \; \{ , v \}) [, \texttt{mutual}]
%  \\[0.5em]
\dimTy ::= \; & \mathbb{N}_{>0} \\
\nonterm{rvar} ::= \; & [\text{\term{a}-\term{z}$\,$\term{A}-\term{Z}$\,$\term{0}-\term{9}}]+
\end{array}
\end{align*}
\caption{Specification syntax (EBNF grammar)}
\label{fig:syntax}
\end{figure}

@@

\begin{figure}
\begin{align*}
\setlength{\arraycolsep}{0.05em}
\begin{array}{c}
\framebox{$\textit{region} \equiv \textit{region}'$} \\[1em]
\begin{array}{rll}
(\textsc{F\; \texttt{+} \;F}) \;\; &
\stenFwdS{n \, \textsf{max} \, m}{d} \\
 \equiv \; & \stenFwdS{n}{d} \; \texttt{+} \; \stenFwdS{m}{d} \\[1em]
%
(\textsc{B\; \texttt{+} \;B}) \;\; &
\stenBwdS{n \, \textsf{max} \, m}{d} \\
 \equiv \; & \stenBwdS{n}{d} \; \texttt{+} \; \stenBwdS{m}{d} \\[1em]
%
(\textsc{C\; \texttt{+} \;C}) \;\; &
\stenCenS{n \, \textsf{max} \, m}{d} \\
\equiv \; & \stenCenS{n}{d} \; \texttt{+} \; \stenCenS{m}{d} \\[1em]
%
(\textsc{C\; \texttt{+} \;F}) \;\; & \stenCenS{n}{d} \\
\textit{$m \leq n$} \; \equiv \; & \stenCenS{n}{d} \; \texttt{+} \;
                      \stenFwdS{m}{d} \\[1em]
%
(\textsc{C\; \texttt{+} \;B}) \;\; &
\stenCenS{n}{d} \\
\textit{$m \leq n$} \; \equiv \; & \stenCenS{n}{d} \; \texttt{+} \;
                      \stenBwdS{m}{d} \\[1em]
(\textsc{B\; \texttt{+} \;F}) \;\; &
\stenCenS{n}{d} \\
\equiv \; & \stenFwdS{n}{d} \; \texttt{+} \; \stenBwdS{n}{d} \\[1em]
%
(\texttt{+}\textsc{idem}) \;\; & S \; \texttt{+} \; S \; \equiv \; S \\[0.5em]
%
(\texttt{+}\textsc{comm}) \;\; & S \; \texttt{+} \; T \; \equiv \; T \;
                       \texttt{+} \; S \\[0.5em]
%
(\texttt{+}\textsc{assoc}) \;\; & R \; \texttt{+} \; (S \; \texttt{+} \; T) \; \equiv \; (R \;
                       \texttt{+} \; S) \; \texttt{+} \; T \\[0.5em]
(\texttt{*}\textsc{comm}) \;\; & S \; \texttt{*} \; T \; \equiv \; T \;
                       \texttt{*} \; S \\[0.5em]
%
(\texttt{*}\textsc{assoc}) \;\; & R \; \texttt{*} \; (S \; \texttt{*} \; T) \; \equiv \; (R \;
                       \texttt{*} \; S) \; \texttt{*} \; T \\[0.5em]
(\textsc{dist}) \;\; & R \; \texttt{*} \; (S \; \texttt{+} \; T) \; \equiv \; (R \;
                       \texttt{*} \; S) \; \texttt{+} \; (R
                       \; \texttt{*} \; T) \\[0.5em]
\end{array} \\ \\
%
\framebox{${\textit{specDec} :: \texttt{v}} \equiv
{\textit{specDec'} :: \texttt{v}}$} \\[1em]
\begin{array}{rl}
% MUTUAL
%(\textsc{mutual}) \; &
%\texttt{stencil} \; \texttt{dependency} (\texttt{v}), \texttt{mutual}
%  :: \texttt{w}
%\\
%\equiv \; & \texttt{stencil} \; \texttt{dependency} (\texttt{w}), \texttt{mutual} ::
%  \texttt{v} \\[0.5em]
%(\textsc{coalesce}) \; &
%\texttt{stencil} \; \texttt{dependency} (\bar{v})
%  :: \texttt{v} \\
%& ; \texttt{stencil} \; \texttt{dependency} (\bar{w})
%  :: \texttt{v}
%\\
%\equiv \; & \texttt{stencil} \; \texttt{dependency} (\bar{v}, \bar{w}) ::
%  \texttt{v} \\[0.5em]
(\textsc{spatial}) \; &
(\textit{region} \equiv \textit{region'}) \\
\Rightarrow \; & (\texttt{stencil} \; [\textit{mod}] \; [\textit{approxMod},] \;
\textit{region} \\
& \!\!\equiv \; \texttt{stencil} \; [\textit{mod}] \;
            [\textit{approxMod},] \;
\textit{region}') \\
\end{array}
\end{array}
\end{align*}
\caption{Equations on specifications}
\label{fig:equations}
\end{figure}

\paragraph{Equations}
Figure~\ref{fig:equations} lists the equational theory for
specifications. This is broken down into equations on regions only,
given by the equivalence relation $\textit{region} \equiv \textit{region}'$ and
on specification declarations assigned to variables via the equivalence relation
$\textit{specDec} :: \texttt{v} \equiv
\textit{specDec'} :: \texttt{v}$.

To save space we use abbreviations
\term{refl}, \term{fwd}, \term{bwd}, and \term{cen} for \term{reflexive},
\term{forward}, \term{backward}, and \term{centered} regions respectively.
The equations reveal something of their semantics.  The first six
equations (with labels of the form $\ldots \texttt{+} \ldots$)
explain the behaviour of overlapping regions with sum
\term{+}. The next six equations explain the algebraic behaviour of
\term{+} and \term{*}.  Together, we see that \term{+} is an
associative, commutative, idempotent binary operator that distributes
with \term{*}, which is associative and commutative. This distribution
of \term{*} over \term{+} is key as it is used internally to give a
normal form for stencil specifications akin to Disjunctive Normal Form
(DNF) (where \term{+} is taken as ``disjunction'' and \term{*} as the
``conjunction''). We revisit this normalisation in Section~\ref{}.

@@

Figure~\ref{fig:inequations}

@@

We provide a semantic model of the stencil specification language as
the basis for building correct implementations, showing the soundness
of the equational theory and approximations, and as a basis for the
inference, checking, and program synthesis algorithms.
The semantics of stencil specifications can be given a denotational
model as sets of vectors, represent array indexing.

\begin{definition}
  A specification of maximum dimensionality $n$, is modelled by a
  set of column vectors of size $n$  ($n$-dimensional vectors /
  $n$-vectors for short)
  with values drawn from $\mathbb{Z} \cup \{\infty\}$, representing
  relative indices. The $\infty$ element represents non-relative
  indices (\eg{}, constants), which are seen as infinite with respect
  to the frame of reference of a relative offset from an induction
  variable.

  For example, the array index $(i, j+1)$ corresponds to the vector
  $\vtwoh{0}{1}$, whereas, the $(0, j)$ corresponds to the
  vector $\vtwoh{\infty}{0}$ since $0$ is a non-relative index.
\end{definition}

Figure~\ref{fig:region-model} outlines the core rules for regions
via the interpretation function $\interp{-}$.  We explain the
intermediate notations and definitions:

\begin{definition}A \emph{single-entry vector} of size $n$, denoted
$\singleEntry{r}{n}$, is a vector
where the $r^{th}$ entry is $1$ and all others are $\infty$, \eg{},
$\singleEntry{1}{2} = \vtwoh{1}{\infty}$.
\end{definition}

\begin{definition}A \emph{zero-entry vector} of size $n$, denoted
$\zeroEntry{r}{n}$, is a vector where the $r^{th}$ entry is $0$ and all others
are $\infty$, \eg{}, $\zeroEntry{1}{2} = \vtwoh{0}{\infty}$.
\end{definition}

@@

\begin{figure}
\input{region-model.tex}
%\textit{$a^m \in A$ means there are $n$ copies of $a$ in
%  the multi-set $A$}
\caption{Set model of regions}
\label{fig:region-model}
\end{figure}

\begin{figure}
\input{spatial-model.tex}
\caption{Model of specifications}
\label{fig:spatial-model}
\end{figure}

@@

For each span $S = [\vect{l} \; \vect{u}]$,
for each dimension, each pair of entries in the lower- and upper-bound
vectors is mapped to a concrete \textit{spec}, as defined by the
$\mathsf{conv}$ function:
%%
\begin{align*}
& \mathsf{conv}_i\; [\vect{l} \; \vect{u}] = \\
& \begin{cases}
%% FWD
 \stenFwd{\vect{u}_i}{i}  & \vect{l}_i = 0 \; \wedge \; \vect{u}_i > 0 \\[0.7em]
 \texttt{irreflexive(dims=$i$)}, & \\
 \qquad \stenFwd{\vect{u}_i}{i}  & \vect{l}_i = 1 \; \wedge \; \vect{u}_i > 0 \\[0.7em]
%% BWD
 \stenBwd{|\vect{l}_i|}{i}  & \vect{l}_i < 0 \; \wedge \; \vect{u}_i = 0 \\[0.7em]
 \texttt{irreflexive(dims=$i$)}, & \\
 \quad\;\; \stenBwd{|\vect{l}_i|}{i}  & \vect{l}_i < 0 \; \wedge \; \vect{u}_i = -1 \\[0.7em]
%% CEN
 \stenCen{\vect{u}_i}{i}  &\vect{l}_i < 0 \; \wedge \; \vect{u}_i > 0
 \; \\
 & \qquad\quad \wedge \; |\vect{l}_i| = \vect{u}_i \\[0.7em]
 \stenBwd{|\vect{l}_i|}{i} & \vect{l}_i < 0 \; \wedge \; \vect{u}_i >
 0 \\
\quad \texttt{+} \; \stenFwd{\vect{u}_i}{i}  & \qquad\quad \wedge \; |\vect{l}_i| \neq \vect{u}_i \\[0.7em]
% ABSOLUTE REP
 \texttt{irreflexive(dims=$i$)} & \vect{l}_i = \infty \; \vee \; \vect{u}_i
 = \infty \\[0.7em]
% BOUND
 \texttt{atMost}, \stenFwd{\vect{u}_i}{i} & \vect{l}_i > 1 \\[0.7em]
 \texttt{atMost}, \stenBwd{|\vect{l}_i|}{i} \quad & \vect{l}_i < (-1) \\[0.7em]
  \end{cases}
\end{align*}
%%
The first two cases map to \texttt{forward} stencils since the lower
bound is $0$ or $1$, with the depth determined by the upper bound of
the span. If the lower bound is $1$, then the
$0$-offset point is missing so the specification is given an
\texttt{irreflexive} modifier. The second two cases are similar for
\texttt{backward} but now the depth of the region is determined by the
absolute of the lower bound. In the fifth-case, when the lower bound
is negative and the upper bound is positive and they have the same
magnitude, then a \texttt{centered} region is constructed with
the depth as the common magnitude. If the magnitudes are not the same
(sixth case) then separated \texttt{forward} and \texttt{backward}
regions are summed, with their corresponding depths.

>>>>

@@

\newcommand{\stenRefl}[1]{\term{reflexive} \, (\term{dim=}#1)}
\newcommand{\stenFwd}[3]{\term{forward} \, (\term{depth=}#1,
  \term{dim=}#2{#3})}
\newcommand{\stenBwd}[3]{\term{backward} \, (\term{depth=}#1,
  \term{dim=}#2{#3})}
\newcommand{\stenCen}[3]{\term{centered} \, (\term{depth=}#1,
  \term{dim=}#2{#3})}
\newcommand{\irrefl}{\texttt{irreflexive}}

\newcommand{\stenReflS}[1]{\term{refl} \, (\term{dim=}#1)}
\newcommand{\stenFwdS}[2]{\term{fwd} \, (\term{depth=}#1,
  \term{dim=}#2)}
\newcommand{\stenBwdS}[2]{\term{bwd} \, (\term{depth=}#1,
  \term{dim=}#2)}
\newcommand{\stenCenS}[2]{\term{cen} \, (\term{depth=}#1,
  \term{dim=}#2)}
\newcommand{\irreflS}{\texttt{irrefl}}

\newcommand{\stenFwdSR}[3]{\term{fwd} (\term{depth=}#1,
  \term{dim=}#2, #3)}
\newcommand{\stenBwdSR}[3]{\term{bwd} (\term{depth=}#1,
  \term{dim=}#2, #3)}
\newcommand{\stenCenSR}[3]{\term{cen} (\term{depth=}#1,
  \term{dim=}#2, #3)}
\newcommand{\stenReflSR}[1]{\term{refl} (\term{dim=}#1)}

@@

\begin{figure}[t]
\begin{align*}
\def\arraystretch{1.2}
\setlength{\arraycolsep}{0.2em}
\newcommand{\dimTy}{\mathbb{D}}
\begin{array}{rl}
\nonterm{specification} ::= & \nonterm{regionDec} \mid \nonterm{specDec} \\
\nonterm{specDec} ::= & \term{stencil} \; \nonterm{specTop} \;
                        \texttt{::} \; v
  \\
\nonterm{regionDec} ::= &  \texttt{region} \; \nonterm{rvar} \; \texttt{=} \;
                         \nonterm{region}\\[1em]
%\nonterm{spec} ::= & \nonterm{spatial} \mid \nonterm{temporal}
%\\[1em]
\nonterm{specTop} ::= & [\nonterm{approxMod},] \; \nonterm{spec} \\
\nonterm{spec} ::= & [\nonterm{multMod},] \; \nonterm{region} \\
\nonterm{multMod} ::= & \term{readOnce} \\
\nonterm{approxMod} ::= & \term{atMost} \; \mid \; \term{atLeast} \\[0.1em]
\nonterm{region} ::= 
        & \stenFwd{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{irreflexive}]} \\
\mid \; & \stenBwd{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{irreflexive}]} \\
\mid \; & \stenCen{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{irreflexive}]} \\
\mid \; & \stenRefl{\dimTy}  \\
\mid \; & \nonterm{region} \; \term{+} \; \nonterm{region} \\
\mid \; & \nonterm{region} \; \term{*} \; \nonterm{region} \\
\mid \; & \nonterm{rvar}  \\[0.5em]
%\nonterm{temporal} ::= \; & \term{dependency} \; (v \; \{ , v \}) [, \texttt{mutual}]
%  \\[0.5em]
\dimTy ::= \; & \mathbb{N}_{>0} \\
\nonterm{rvar} ::= \; & [\text{\term{a}-\term{z}$\,$\term{A}-\term{Z}$\,$\term{0}-\term{9}}]+
\end{array}
\end{align*}
\caption{Specification syntax (EBNF grammar)}
\label{fig:syntax}
\end{figure}

@@

\begin{figure}
\begin{align*}
\setlength{\arraycolsep}{0.05em}
\begin{array}{c}
\framebox{$\textit{region} \equiv \textit{region}'$} \\[1em]
\begin{array}{rll}
(\textsc{F\, \texttt{+} \,F}) \;\; &
\stenFwdSR{n \, \textsf{max} \, m}{d}{r \vee s} \\
 \equiv \; & \stenFwdSR{n}{d}{r} \; \texttt{+} \; \stenFwdSR{m}{d}{s} \\[1em]
%
%
(\textsc{B\, \texttt{+} \,B}) \;\; &
\stenBwdSR{n \, \textsf{max} \, m}{d}{r \vee s} \\
 \equiv \; & \stenBwdS{n}{d}{r} \; \texttt{+} \; \stenBwdSR{m}{d}{s} \\[1em]
%
%
(\textsc{C\, \texttt{+} \,C}) \;\; &
\stenCenSR{n \, \textsf{max} \, m}{d}{r \vee s} \\
\equiv \; & \stenCenSR{n}{d}{r} \; \texttt{+} \; \stenCenSR{m}{d}{s} \\[1em]
%
%
(\textsc{C\, \texttt{+} \,F}) \;\; & \stenCenSR{n}{d}{r \vee s} \\
\textit{$m \leq n$} \; \equiv \; & \stenCenSR{n}{d}{r} \; \texttt{+} \;
                      \stenFwdSR{m}{d}{s} \\[1em]
%
%
(\textsc{C\, \texttt{+} \,B}) \;\; &
\stenCenSR{n}{d}{r \vee s} \\
\textit{$m \leq n$} \; \equiv \; & \stenCenSR{n}{d}{r} \; \texttt{+} \;
                      \stenBwdSR{m}{d}{s} \\[1em]
%
%
(\textsc{B\, \texttt{+} \,F}) \;\; &
\stenCenSR{n}{d}{r \vee s} \\
\equiv \; & \stenFwdSR{n}{d}{r} \; \texttt{+} \; \stenBwdSR{n}{d}{s}
  \\[1em]
%
%
(\textsc{R\, \texttt{+} \,F}) \;\; &
\stenFwdS{n}{d} \\
\equiv \; & \stenReflSR{d} \; \texttt{+} \; \stenFwdSR{n}{d}{r} \\[1em]
%
%
(\textsc{R\, \texttt{+} \,B}) \;\; &
\stenBwdS{n}{d} \\
\equiv \; & \stenReflSR{d} \; \texttt{+} \; \stenBwdSR{n}{d}{r} \\[1em]
%
%
(\textsc{R\, \texttt{+} \,C}) \;\; &
\stenCenS{n}{d} \\
\equiv \; & \stenReflSR{d} \; \texttt{+} \; \stenCenSR{n}{d}{r} \\[1em]
%
%
(\textsc{irrefl}) \;\; & (R \, \term{*} \, S^\irreflS) \, \term{+} \,
                         (R^\irreflS \, \term{*} \, S) \; 
\equiv \; R \, \term{*} \, S \\[0.5em]
%
%
(\texttt{+}\textsc{idem}) \;\; & S \; \texttt{+} \; S \; \equiv \; S \\[0.5em]
%
(\texttt{+}\textsc{comm}) \;\; & S \; \texttt{+} \; T \; \equiv \; T \;
                       \texttt{+} \; S \\[0.5em]
%
(\texttt{+}\textsc{assoc}) \;\; & R \; \texttt{+} \; (S \; \texttt{+} \; T) \; \equiv \; (R \;
                       \texttt{+} \; S) \; \texttt{+} \; T \\[0.5em]
(\texttt{*}\textsc{comm}) \;\; & S \; \texttt{*} \; T \; \equiv \; T \;
                       \texttt{*} \; S \\[0.5em]
%
(\texttt{*}\textsc{assoc}) \;\; & R \; \texttt{*} \; (S \; \texttt{*} \; T) \; \equiv \; (R \;
                       \texttt{*} \; S) \; \texttt{*} \; T \\[0.5em]
(\textsc{dist}) \;\; & R \; \texttt{*} \; (S \; \texttt{+} \; T) \; \equiv \; (R \;
                       \texttt{*} \; S) \; \texttt{+} \; (R
                       \; \texttt{*} \; T)
\end{array}
%
%\framebox{${\textit{specDec} :: \texttt{v}} \equiv
%{\textit{specDec'} :: \texttt{v}}$} \\[1em]
%\begin{array}{rl}
% MUTUAL
%(\textsc{mutual}) \; &
%\texttt{stencil} \; \texttt{dependency} (\texttt{v}), \texttt{mutual}
%  :: \texttt{w}
%\\
%\equiv \; & \texttt{stencil} \; \texttt{dependency} (\texttt{w}), \texttt{mutual} ::
%  \texttt{v} \\[0.5em]
%(\textsc{coalesce}) \; &
%\texttt{stencil} \; \texttt{dependency} (\bar{v})
%  :: \texttt{v} \\
%& ; \texttt{stencil} \; \texttt{dependency} (\bar{w})
%  :: \texttt{v}
%\\
%\equiv \; & \texttt{stencil} \; \texttt{dependency} (\bar{v}, \bar{w}) ::
%  \texttt{v} \\[0.5em]
%\end{array}
\end{array}
\end{align*}
\caption{Equations on specifications}
\label{fig:equations}
\end{figure}

Figure~\ref{fig:equations} lists the equational theory for
specifications which is defined by the binary relation $\equiv$
between region terms. Stencil declarations are considered
equal when they have the same modifiers and $\equiv$ equates
their regions, \ie{}:
\begin{align*}
& \textit{region} \equiv \textit{region'} \\
& \quad \Rightarrow \;\; \texttt{stencil} \; [\textit{approxMod},] \; [\textit{multMod},] \;
\textit{region} \\[-0.4em]
& \quad\quad \;\; \equiv \;\texttt{stencil} \; [\textit{approxMod},] \;
            [\textit{multMod},] \; \textit{region}'
\end{align*}
%
To save space in the equations of Figure~\ref{fig:equations}, we use abbreviations
\term{refl}, \term{fwd}, \term{bwd}, and \term{cen} for \term{reflexive},
\term{forward}, \term{backward}, and \term{centered} regions
respectively. Futhermore, the abbreviations include syntactic sugar
for quantifying over the abseence or presence of the \irrefl{}
attribute on region constants as a boolean. For example, 
the syntactic sugar in our equations for \term{fwd} is defined:
%%
\begin{align*}
& \stenFwdSR{n}{d}{r} 
\; := \; \\[-0.4em]
& \qquad \begin{cases}
\stenFwd{n}{d}{, \irrefl} & \textit{if} \;\, \neg \, r \\
\stenFwd{n}{d}{}  & \textit{if} \;\,  r
\end{cases}
\end{align*}
%
and similarly for \term{bwd} and \term{cen}.
We also frequently abbreviate \irrefl{} to \irreflS{}.

The equations reveal a partial semantics for stencil
specifications, which we briefly summarise.

\paragraph{Overlapping regions with +}
The first nine
equations (with labels of the form $\ldots \texttt{+} \ldots$)
explain the notion of \emph{overlapping} regions with sum
\term{+}. Note that in the first six, $r$ and $s$ range 
over booleans denoting the absence (true) or presence (false)
of \irrefl{} which are combined via disjunction. For example,
the rule $(\textsc{B} \, \term{+} \,
\textsc{F})$ explains that the addition of \term{backward}
and \term{forward} regions in the same dimension and the same
depth is equivalent to a \term{centered} region of that depth.
If the \term{forward} part is irrefleixve (\ie{}, $r =
\textit{false}$) but \term{backward} is not (\ie{}, $s =
\textit{true}$) then their addition is equal to a \term{centered}
region with no \irrefl{} modifier by the disjunction ($r \vee s =
\textit{true}$), \ie{}:
%%
\begin{align*}
& \stenCenS{n}{d}  \\[-0.2em]
\equiv \;\; & \stenFwdSR{n}{d}{\irreflS} \; \texttt{+} \; \stenBwdS{n}{d}
\end{align*}
%%
That is, the \term{fwd} region is irreflexive describing code which is missing
a zero-offset neighbourhood index (\eg{}, missing $\texttt{a(i)}$), but the addition of
the \term{bwd} has no irreflexive attribute and thus provides the
missing zero offset index.

The equation (\textsc{irrefl}) explains an annihilation of 
\irrefl{} using a further syntactic sugar.
We write $R^{\irreflS}$ to denote a region constant with an
irreflexive attribute, \eg{}, 
%%
\begin{align*}
\stenFwdS{n}{d}^{\irreflS} \, := \,
\stenFwdSR{n}{d}{\irreflS}
\end{align*}
%% 
Thus, (\textsc{irrefl}) explains the sum of a 
product of two regions $R$ and $S$ with itself,
where in each component of the sum one of $R$
or $S$ has the irreflexive attribute. These cancel to give
just the specification $R \, \term{*} \, S$.

\paragraph{Interaction between \term{+} and \term{*}}

The last six equations (from (\textsc{\term{+}idem}) inclusive), explain
the algebraic behaviour of \term{+} and \term{*}.  Together, we see
that \term{+} is an associative, commutative, idempotent binary
operator that distributes with \term{*}, which is associative and
commutative. This distribution of \term{*} over \term{+} is key as it
is used internally to give a normal form for stencil specifications
akin to Disjunctive Normal Form (DNF) (where \term{+} is taken as
``disjunction'' and \term{*} as the ``conjunction''). We revisit this
normalisation in Section~\ref{}.

@@

Figure~\ref{fig:inequations}\dnote{TODO: this might
need to include notions of consistency}

@@

We define a semantic model of our stencil specification language as
the basis for building correct implementations, showing the soundness
of the equational theory and approximations, and as a basis for the
inference, checking, and program synthesis algorithms.
The semantics of stencil specifications can be given a denotational
model as sets of vectors, represent array indexing.

\begin{definition}
  A specification of maximum dimensionality $n$, is modelled by a
  set of vectors of size $n$ ($n$-dimensional vectors or 
  $n$-vectors for short)
  with values drawn from $\mathbb{Z} \cup \{\infty\}$, representing
  relative indices. The element $\infty$ represents non-relative
  indices, such as constants, which are seen as infinite with respect
  to the frame of reference of a relative offset from an induction
  variable.

  For example, the array index term $(i, j+1)$ corresponds to the vector
  $\vtwoh{0}{1}$, whereas, the index term $(3, j)$ corresponds to the
  vector $\vtwoh{\infty}{0}$ since $3$ is a non-relative index.
\end{definition}

We define the interpretation function $\inter{-}$ overloaded on regions
in Figure~\ref{fig:region-model} and on the top-level of 
a specification \textit{specTop} in
Figure~\ref{fig:spatial-model}. Some
intermediate notations and definitions are used.

\begin{definition}A \emph{single-entry vector} of size $n$, denoted
$\singleEntry{r}{n}$, is a vector
where the $r^{th}$ entry is $1$ and all others are $\infty$. For
example, 
$\singleEntry{1}{2} = \vtwoh{\infty}{1}$ and $\singleEntry{2}{3} = 
\vthreeh{1}{\infty}{1}$. 
\end{definition}

\begin{definition}A \emph{zero-entry vector} of size $n$, denoted
$\zeroEntry{r}{n}$, is a vector where the $r^{th}$ entry is $0$ and all others
are $\infty$, \eg{}, $\zeroEntry{1}{2} = \vtwoh{\infty}{0}$.
\end{definition}

@@

\begin{figure}
\input{region-model.tex}
%\textit{$a^m \in A$ means there are $n$ copies of $a$ in
%  the multi-set $A$}
\caption{Model of regions, $\interp{\textit{region}}$}
\label{fig:region-model}
\end{figure}

\begin{figure}
\input{spatial-model.tex}
\caption{Model of specifications, $\interp{\textit{specTop}}$}
\label{fig:spatial-model}
\end{figure}

@@

For each span $S = [\vect{l} \; \vect{u}]$,
for each dimension, each pair of entries in the lower- and upper-bound
vectors is mapped to a concrete \textit{spec}, as defined by the
$\mathsf{conv}$ function:
%%
\begin{align*}
& \mathsf{conv}_i\; [\vect{l} \; \vect{u}] = \\
& \begin{cases}
%% FWD
 \stenFwd{\vect{u}_i}{i}{}  & \vect{l}_i = 0 \; \wedge \; \vect{u}_i > 0 \\[0.7em]
 \stenFwd{\vect{u}_i}{i}{, \irreflS}  & \vect{l}_i = 1 \; \wedge \; \vect{u}_i > 0 \\[0.7em]
%% BWD
 \stenBwd{|\vect{l}_i|}{i}{}  & \vect{l}_i < 0 \; \wedge \; \vect{u}_i = 0 \\[0.7em]
\stenBwd{|\vect{l}_i|}{i}{, \irreflS} & \vect{l}_i < 0 \; \wedge \; \vect{u}_i = -1 \\[0.7em]
%% CEN
 \stenCen{\vect{u}_i}{i}{}  &\vect{l}_i < 0 \; \wedge \; \vect{u}_i > 0
 \; \\
 & \qquad\quad \wedge \; |\vect{l}_i| = \vect{u}_i \\[0.7em]
 \stenBwd{|\vect{l}_i|}{i}{} & \vect{l}_i < 0 \; \wedge \; \vect{u}_i >
 0 \\
\quad \texttt{+} \; \stenFwd{\vect{u}_i}{i}{}  & \qquad\quad \wedge \; |\vect{l}_i| \neq \vect{u}_i \\[0.7em]
% ABSOLUTE REP
 \epsilon & \vect{l}_i = \infty \; \vee \; \vect{u}_i
 = \infty \\[0.7em]
% BOUND
 \texttt{atMost}, \stenFwd{\vect{u}_i}{i}{} & \vect{l}_i > 1 \\[0.7em]
 \texttt{atMost}, \stenBwd{|\vect{l}_i|}{i}{} \quad & \vect{l}_i < (-1) \\[0.7em]
  \end{cases}
\end{align*}
%%
The first two cases map to \texttt{forward} stencils since the lower
bound is $0$ or $1$, with the depth determined by the upper bound of
the span. If the lower bound is $1$, then the
$0$-offset point is missing so the specification is given an
\texttt{irreflexive} modifier. The second two cases are similar for
\texttt{backward} but now the depth of the region is determined by the
absolute of the lower bound. In the fifth-case, when the lower bound
is negative and the upper bound is positive and they have the same
magnitude, then a \texttt{centered} region is constructed with
the depth as the common magnitude. If the magnitudes are not the same
(sixth case) then separated \texttt{forward} and \texttt{backward}
regions are summed, with their corresponding depths.
